---
title: "Bayesian Fit Demo Workflow"
output: html_document
---

# Overview

This notebook combines the Bayesian helper functions, training pipeline, and
prediction flow into one place.

The workflow is:

1. Load the engineered training table created by
   `scripts/export_bayesian_demo_dataset.py`.
2. Fit a manual Bayesian probit model with Gibbs sampling.
3. Save the posterior draws and coefficient summary.
4. Score a single user for a selected brand size using their measurements.

The original command-line arguments from the standalone scripts are replaced
here with editable variables in the setup chunk below.

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# Find the project root whether the notebook is run from the repo root
# or from inside the scripts/ directory.
detect_project_root <- function() {
  candidates <- unique(c(
    normalizePath(getwd(), mustWork = FALSE),
    normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
  ))

  for (candidate in candidates) {
    has_data <- dir.exists(file.path(candidate, "Data"))
    has_scripts <- dir.exists(file.path(candidate, "scripts"))
    if (has_data && has_scripts) {
      return(candidate)
    }
  }

  stop("Could not locate the project root from the current working directory.")
}

project_root <- detect_project_root()
data_dir <- file.path(project_root, "Data")
artifacts_dir <- file.path(project_root, "artifacts")

training_csv <- file.path(data_dir, "bayesian_demo_training.csv")
brand_csv <- file.path(data_dir, "demo_brand_sizing.csv")
model_path <- file.path(artifacts_dir, "bayesian_probit_model.rds")
coefficient_summary_path <- file.path(
  artifacts_dir,
  "bayesian_probit_coefficients.csv"
)

# Training hyperparameters.
burn_in <- 1000L
samples <- 2000L
thin <- 2L
prior_var <- 25
seed <- 42L

# Example request for the demo prediction step.
# Edit these values to test a different user and selected size.
prediction_input <- list(
  garment_type = "tops",
  size_label = "M",
  height_inches = 65,
  bust = 36,
  waist = 28,
  hips = 38
)

set.seed(seed)
dir.create(artifacts_dir, recursive = TRUE, showWarnings = FALSE)
```

# Helper Functions

These helpers come from `bayesian_probit_utils.R`. They are kept inline here so
the notebook is fully self-contained.

```{r helper-functions}
GARMENT_LEVELS <- c("tops", "bottoms", "dresses")
SIZE_LEVELS <- c("XS", "S", "M", "L", "XL")
CONTINUOUS_FEATURES <- c(
  "size_ord",
  "height_gap",
  "bust_gap",
  "waist_gap",
  "hips_gap",
  "abs_height_gap",
  "abs_bust_gap",
  "abs_waist_gap",
  "abs_hips_gap"
)

build_design_matrix <- function(df, numeric_stats = NULL) {
  df <- as.data.frame(df)
  df$garment_type <- factor(df$garment_type, levels = GARMENT_LEVELS)
  df$size_label <- factor(df$size_label, levels = SIZE_LEVELS, ordered = TRUE)

  # Learn scaling statistics from the training data once, then reuse them
  # unchanged at prediction time.
  if (is.null(numeric_stats)) {
    means <- vapply(df[CONTINUOUS_FEATURES], mean, numeric(1), na.rm = TRUE)
    sds <- vapply(df[CONTINUOUS_FEATURES], sd, numeric(1), na.rm = TRUE)
    sds[is.na(sds) | sds == 0] <- 1
    numeric_stats <- list(means = means, sds = sds)
  }

  for (feature_name in CONTINUOUS_FEATURES) {
    df[[feature_name]] <- (
      df[[feature_name]] - numeric_stats$means[[feature_name]]
    ) / numeric_stats$sds[[feature_name]]
  }

  design_formula <- stats::as.formula(
    "~ garment_type + size_label + size_ord + height_gap + bust_gap + waist_gap + hips_gap + abs_height_gap + abs_bust_gap + abs_waist_gap + abs_hips_gap + height_missing + bust_missing + waist_missing + hips_missing"
  )

  X <- stats::model.matrix(design_formula, data = df)
  y <- if ("is_fit" %in% names(df)) as.integer(df$is_fit) else NULL

  list(
    X = X,
    y = y,
    numeric_stats = numeric_stats,
    feature_names = colnames(X)
  )
}

sample_latent_z <- function(y, linear_predictor) {
  y <- as.integer(y)
  latent <- numeric(length(y))

  positive_idx <- which(y == 1L)
  if (length(positive_idx) > 0) {
    lower_cdf <- stats::pnorm(
      0,
      mean = linear_predictor[positive_idx],
      sd = 1
    )
    u <- lower_cdf + stats::runif(length(positive_idx)) * (1 - lower_cdf)
    u <- pmin(pmax(u, 1e-10), 1 - 1e-10)
    latent[positive_idx] <- stats::qnorm(
      u,
      mean = linear_predictor[positive_idx],
      sd = 1
    )
  }

  negative_idx <- which(y == 0L)
  if (length(negative_idx) > 0) {
    upper_cdf <- stats::pnorm(
      0,
      mean = linear_predictor[negative_idx],
      sd = 1
    )
    u <- stats::runif(length(negative_idx)) * upper_cdf
    u <- pmin(pmax(u, 1e-10), 1 - 1e-10)
    latent[negative_idx] <- stats::qnorm(
      u,
      mean = linear_predictor[negative_idx],
      sd = 1
    )
  }

  latent
}

rmvnorm_chol <- function(mean_vec, chol_cov) {
  as.numeric(mean_vec + t(chol_cov) %*% stats::rnorm(length(mean_vec)))
}

fit_bayesian_probit <- function(
  X,
  y,
  burn_in = 1000L,
  samples = 2000L,
  thin = 2L,
  prior_var = 25,
  progress_every = 500L
) {
  X <- as.matrix(X)
  y <- as.integer(y)
  p <- ncol(X)

  # Albert-Chib style Gaussian prior and Gaussian conditional update.
  prior_precision <- diag(1 / prior_var, p)
  posterior_cov <- solve(crossprod(X) + prior_precision)
  posterior_chol <- chol(posterior_cov)

  total_iterations <- burn_in + samples * thin
  beta <- rep(0, p)
  beta_draws <- matrix(NA_real_, nrow = samples, ncol = p)
  colnames(beta_draws) <- colnames(X)
  draw_idx <- 1L

  for (iteration in seq_len(total_iterations)) {
    linear_predictor <- as.numeric(X %*% beta)
    latent_z <- sample_latent_z(y, linear_predictor)

    posterior_mean <- posterior_cov %*% crossprod(X, latent_z)
    beta <- rmvnorm_chol(posterior_mean, posterior_chol)

    if (iteration > burn_in && ((iteration - burn_in) %% thin == 0L)) {
      beta_draws[draw_idx, ] <- beta
      draw_idx <- draw_idx + 1L
    }

    if (progress_every > 0L && (iteration %% progress_every == 0L)) {
      cat(sprintf("Iteration %d / %d\n", iteration, total_iterations))
    }
  }

  list(
    beta_draws = beta_draws,
    posterior_mean = colMeans(beta_draws),
    posterior_sd = apply(beta_draws, 2, stats::sd)
  )
}

posterior_fit_summary <- function(X_new, beta_draws) {
  X_new <- as.matrix(X_new)
  linear_predictor <- as.numeric(X_new %*% t(beta_draws))
  fit_probs <- stats::pnorm(linear_predictor)

  c(
    mean = mean(fit_probs),
    lower = unname(stats::quantile(fit_probs, probs = 0.05)),
    upper = unname(stats::quantile(fit_probs, probs = 0.95))
  )
}

create_demo_feature_row <- function(
  brand_chart,
  garment_type,
  size_label,
  height_inches = NA_real_,
  bust = NA_real_,
  waist = NA_real_,
  hips = NA_real_
) {
  chart_row <- brand_chart[
    brand_chart$garment_type == garment_type &
      brand_chart$size_label == size_label,
    ,
    drop = FALSE
  ]

  if (nrow(chart_row) != 1L) {
    stop(
      sprintf(
        "Could not find a unique chart row for garment_type=%s and size_label=%s",
        garment_type,
        size_label
      ),
      call. = FALSE
    )
  }

  centers <- list(
    height = (chart_row$height_min_inches + chart_row$height_max_inches) / 2,
    bust = (chart_row$bust_min + chart_row$bust_max) / 2,
    waist = (chart_row$waist_min + chart_row$waist_max) / 2,
    hips = (chart_row$hips_min + chart_row$hips_max) / 2
  )

  # If a measurement is missing, we back-fill with the size-chart center and
  # track the missingness explicitly with indicator features.
  observed_height <- if (is.na(height_inches)) centers$height else height_inches
  observed_bust <- if (is.na(bust)) centers$bust else bust
  observed_waist <- if (is.na(waist)) centers$waist else waist
  observed_hips <- if (is.na(hips)) centers$hips else hips

  data.frame(
    garment_type = garment_type,
    size_label = size_label,
    size_ord = match(size_label, SIZE_LEVELS),
    height_gap = observed_height - centers$height,
    bust_gap = observed_bust - centers$bust,
    waist_gap = observed_waist - centers$waist,
    hips_gap = observed_hips - centers$hips,
    abs_height_gap = abs(observed_height - centers$height),
    abs_bust_gap = abs(observed_bust - centers$bust),
    abs_waist_gap = abs(observed_waist - centers$waist),
    abs_hips_gap = abs(observed_hips - centers$hips),
    height_missing = as.integer(is.na(height_inches)),
    bust_missing = as.integer(is.na(bust)),
    waist_missing = as.integer(is.na(waist)),
    hips_missing = as.integer(is.na(hips)),
    stringsAsFactors = FALSE
  )
}
```

# Train the Bayesian Probit Model

This section matches the logic from `train_bayesian_probit.R`.

```{r train-model}
training_df <- utils::read.csv(training_csv, stringsAsFactors = FALSE)
design <- build_design_matrix(training_df)

fit <- fit_bayesian_probit(
  X = design$X,
  y = design$y,
  burn_in = burn_in,
  samples = samples,
  thin = thin,
  prior_var = prior_var
)

posterior_mean_prob <- stats::pnorm(as.numeric(design$X %*% fit$posterior_mean))
predicted_label <- ifelse(posterior_mean_prob >= 0.5, 1L, 0L)

accuracy <- mean(predicted_label == design$y)
precision <- if (sum(predicted_label == 1L) == 0L) {
  0
} else {
  mean(design$y[predicted_label == 1L] == 1L)
}
recall <- if (sum(design$y == 1L) == 0L) {
  0
} else {
  mean(predicted_label[design$y == 1L] == 1L)
}

coefficient_summary <- data.frame(
  feature = colnames(design$X),
  posterior_mean = fit$posterior_mean,
  posterior_sd = fit$posterior_sd,
  stringsAsFactors = FALSE
)

model_payload <- list(
  beta_draws = fit$beta_draws,
  posterior_mean = fit$posterior_mean,
  feature_names = colnames(design$X),
  numeric_stats = design$numeric_stats,
  garment_levels = GARMENT_LEVELS,
  size_levels = SIZE_LEVELS,
  continuous_features = CONTINUOUS_FEATURES,
  training_metrics = list(
    accuracy = accuracy,
    precision = precision,
    recall = recall
  ),
  training_rows = nrow(training_df),
  seed = seed,
  prior_var = prior_var
)

saveRDS(model_payload, model_path)
utils::write.csv(coefficient_summary, coefficient_summary_path, row.names = FALSE)

training_summary <- data.frame(
  rows = nrow(training_df),
  features = ncol(design$X),
  posterior_draws = nrow(fit$beta_draws),
  accuracy = accuracy,
  precision = precision,
  recall = recall
)

training_summary
```

```{r coefficient-preview}
head(coefficient_summary[order(-abs(coefficient_summary$posterior_mean)), ], 10)
```

# Predict Fit for a Selected Size

This section matches the logic from `predict_bayesian_fit.R`. Update
`prediction_input` in the setup chunk to test a new user.

```{r predict-fit}
reloaded_model <- readRDS(model_path)
brand_chart <- utils::read.csv(brand_csv, stringsAsFactors = FALSE)

feature_row <- create_demo_feature_row(
  brand_chart = brand_chart,
  garment_type = prediction_input$garment_type,
  size_label = prediction_input$size_label,
  height_inches = prediction_input$height_inches,
  bust = prediction_input$bust,
  waist = prediction_input$waist,
  hips = prediction_input$hips
)

prediction_design <- build_design_matrix(
  feature_row,
  numeric_stats = reloaded_model$numeric_stats
)
fit_summary <- posterior_fit_summary(
  prediction_design$X,
  reloaded_model$beta_draws
)

chart_row <- brand_chart[
  brand_chart$garment_type == prediction_input$garment_type &
    brand_chart$size_label == prediction_input$size_label,
  ,
  drop = FALSE
]

within_chart <- data.frame(
  measurement = c("height", "bust", "waist", "hips"),
  within_range = c(
    is.na(prediction_input$height_inches) || (
      prediction_input$height_inches >= chart_row$height_min_inches &&
        prediction_input$height_inches <= chart_row$height_max_inches
    ),
    is.na(prediction_input$bust) || (
      prediction_input$bust >= chart_row$bust_min &&
        prediction_input$bust <= chart_row$bust_max
    ),
    is.na(prediction_input$waist) || (
      prediction_input$waist >= chart_row$waist_min &&
        prediction_input$waist <= chart_row$waist_max
    ),
    is.na(prediction_input$hips) || (
      prediction_input$hips >= chart_row$hips_min &&
        prediction_input$hips <= chart_row$hips_max
    )
  )
)

decision <- if (fit_summary[["mean"]] >= 0.5) "likely_fit" else "likely_misfit"
confidence <- if (
  fit_summary[["lower"]] > 0.5 || fit_summary[["upper"]] < 0.5
) {
  "high"
} else {
  "moderate"
}

prediction_result <- data.frame(
  garment_type = prediction_input$garment_type,
  selected_size = prediction_input$size_label,
  posterior_fit_probability = fit_summary[["mean"]],
  credible_interval_lower = fit_summary[["lower"]],
  credible_interval_upper = fit_summary[["upper"]],
  decision = decision,
  confidence = confidence,
  stringsAsFactors = FALSE
)

prediction_result
within_chart
feature_row
```

# Notes

- The notebook assumes `Data/bayesian_demo_training.csv` already exists.
- If needed, regenerate it first with `python scripts/export_bayesian_demo_dataset.py`.
- The prediction section intentionally scores the user against the size they
  selected rather than searching across all sizes.
